---
title: "Shared mutations from the nf-resolveome genotyping output"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include = F}
# rmarkdown::render('reports/20250408_plot_shared_muts_from_nf-resolveome.Rmd', output_file = '20250408_plot_shared_muts_from_nf-resolveome.html', output_dir = 'reports/')

# chunk options
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE,
                      dpi = 600,
                      message = FALSE)

# libraries
library(dplyr)
library(magrittr)
library(ggplot2)
source("bin/utils.R")

# dndscv palette
impact_colours <- c("Missense" = "#5f9fa0",
                    "Nonsense" = "#682c8b",
                    "Essential_Splice" = "#8e55a3",
                    "Indels" = "#d4773f",
                    "Synonymous" = "#b3b4b4",
                    "Non-coding" = "#606060")
```

## Metadata

We load the samplesheet and manual inspection results.

```{r load_ss}
man_insp <-
  readr::read_tsv("data/manual_inspection/2024-12-20_PD63118_PTA_BAF_LoH_CellType_Mut_Summary.tsv") %>%
  # filter out bad cells
  filter((!chr_dropout | is.na(chr_dropout)) & (!suspected_doublet | is.na(suspected_doublet))) %>%
  select(-c(run_id, id, seq_type)) %>%
  # fix loh_1p
  mutate(loh_1p = as.character(as.numeric(loh_1p)))
ss <-
  readr::read_csv("data/resolveome/samplesheet_local.csv") %>%
  left_join(man_insp) %>%
  distinct(id, cell_id, loh_1p, celltype_VDJ_recomb, celltype_SHM,
           celltype_bj_expression)
```

We load the genotyping results.

```{r load_geno}
genos <-
  c("dna", "dnahyb") %>%
  purrr::set_names() %>%
  purrr::map(function(seq_type) {
    paste0("out/nf-resolveome/", seq_type, "/PD63118/genotyping/mutations/PD63118_annotated_mutations.tsv") %>%
      readr::read_tsv()
  }) %>%
  bind_rows() %>%
  left_join(ss) %>%
  select(-c(id, mut_vaf)) %>%
  group_by(across(-matches("_depth$"))) %>%
  summarise(across(matches("_depth$"), \(x) sum(x, na.rm = TRUE))) %>%
  ungroup() %>%
  mutate(mut_vaf = mut_depth / total_depth)
```

We generate a heatmap of all mutations.

```{r plot_heatmap, fig.width = 20, fig.height = 20}
p <-
  genos %>%
  plot_vaf_heatmap(show_rownames = FALSE, p_title = "all mutations",
                   annotations = c("loh_1p", "celltype_VDJ_recomb", "celltype_SHM", "celltype_bj_expression"))
print(p)
```

We plot the VAF distribution of all mutations.

```{r plot_vaf_dist}
genos %>%
  filter(mut_vaf > 0) %>%
  ggplot(aes(x = mut_vaf)) +
  geom_histogram() +
  theme_classic()
```

We filter the mutations.

```{r filter_muts}
genos_filter <-
  genos %>%
  filter(mut_vaf > 0.2, mut_depth > 1)
```

We generate a heatmap of filtered mutations.

```{r plot_heatmap_filtered, fig.width = 20, fig.height = 20}
p <-
  genos_filter %>%
  plot_vaf_heatmap(show_rownames = FALSE,
                   p_title = "filtered mutations",
                   annotations = c("loh_1p", "celltype_VDJ_recomb", "celltype_SHM", "celltype_bj_expression"))
print(p)
```

We plot the VAF distribution of filtered mutations.

```{r plot_vaf_dist_filtered}
genos_filter %>%
  # annotate mutations on 1p
  mutate(`on 1p + 1p LOH` = chr == "1" & pos < 123400000 & loh_1p == 1) %>%
  ggplot(aes(x = mut_vaf, fill = `on 1p + 1p LOH`)) +
  geom_histogram() +
  theme_classic() +
  scale_fill_brewer(palette = "Set1")
```

We plot the distribution of mutations across celltypes.

```{r plot_celltype_dist}
genos_filter %>%
  add_count(gene) %>%
  ggplot(aes(x = reorder(gene, -n), fill = impact)) +
  geom_bar() +
  theme_classic() +
  scale_fill_manual(values = impact_colours) +
  facet_grid(celltype_VDJ_recomb ~ .) +
  labs(x = "gene")
```

We generate a heatmap of the filtered, shared mutations.

```{r plot_heatmap_shared, fig.width = 20, fig.height = 20}
p <-
  genos_filter %>%
  filter(mut_depth > 0) %>%
  group_by(chr, pos, ref, alt) %>%
  filter(n() > 1) %>%
  plot_vaf_heatmap(p_title = "filtered shared mutations",
                   annotations = c("loh_1p", "celltype_VDJ_recomb",
                                   "celltype_bj_expression"))
print(p)
```

We generate a heatmap of the nonsynonymous mutations.

```{r plot_nonsym, fig.width = 20, fig.height = 20}
p <-
  genos_filter %>%
  filter(impact != "Synonymous") %>%
  plot_vaf_heatmap("filtered non-synonymous mutations",
                   annotations = c("loh_1p", "celltype_VDJ_recomb",
                                   "celltype_bj_expression"))
print(p)
```

We generate a heatmap of the TNFRSF14 mutations.

```{r plot_heatmap_filtered_TNFRSF14, fig.width = 20, fig.height = 20}
p <-
  genos_filter %>%
  filter(gene == "TNFRSF14") %>%
  mutate(alt = paste(alt, aachange)) %>%
  plot_vaf_heatmap("TNFRSF14 mutations",
                   annotations = c("loh_1p", "celltype_VDJ_recomb",
                                   "celltype_bj_expression"))
print(p)
```

We generate a heatmap of the non-synonymous mutations in the TNFRSF14 gene.

```{r plot_heatmap_filtered_TNFRSF14_nonsym, fig.width = 20, fig.height = 20}
p <-
  genos_filter %>%
  filter(gene == "TNFRSF14", impact != "Synonymous") %>%
  mutate(alt = paste(alt, aachange)) %>%
  plot_vaf_heatmap("TNFRSF14 non-synonymous mutations",
                   annotations = c("loh_1p", "celltype_VDJ_recomb",
                                   "celltype_bj_expression"))
print(p)
```

We plot the mutations per gene.

```{r muts_per_gene}
p_dat <-
  genos_filter %>%
  filter(impact != "Synonymous") %>%
  group_by(gene) %>%
  summarise(n_cells = n_distinct(cell_id)) %>%
  mutate(prop_cells = n_cells / n_mat_b_cells)
p2 <-
  p_dat %>%
  ggplot(aes(x = reorder(gene, -prop_cells), y = prop_cells)) +
  geom_col(fill = "#d6630d") +
  theme_classic() +
  labs(x = "gene", y = "% of mature B cells with a driver mutation") +
  scale_y_continuous(expand = c(0, 0))

# mutations per gene
p1 <-
  genos_filter %>%
  left_join(man_insp) %>%
  add_count(gene, name = "total") %>%
  count(gene, impact, total, celltype_SHM) %>%
  mutate(impact = ifelse(impact == "no-SNV", "Synonymous", impact) %>% factor(levels = rev(names(impact_colours)))) %>%
  left_join(p_dat %>% distinct(gene, prop_cells)) %>%
  ggplot(aes(x = reorder(gene, -prop_cells), y = n, fill = impact)) +
  geom_col() +
  scale_fill_manual(values = impact_colours) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = "n mutations")

# plot
p1 / p2
```