---
title: "First year report plots"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  out_dir: null
---

```{r setup, include = F, message = F, warning = F, echo = F}
# rmarkdown::render('reports/20250402_first_year_report_plots.Rmd', output_file = '20250402_first_year_report_plots.html', output_dir = 'reports', params = list(out_dir = "out/BaseJumper/bj-expression/filter_cells/"))
# params <- list(out_dir = "out/BaseJumper/bj-expression/filter_cells/")

# chunk options
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE,
                      dpi = 300,
                      message = FALSE)

# libraries
library(magrittr)
library(dplyr)
library(readr)
library(ggplot2)
library(patchwork)
source("bin/utils.R")

# celltype palette
pal <- c("B" = "#2f9e77", "mature B" = "#55c552",
         "non-mature B" = "#397037", "non-lymphocyte" = "#d95f02",
         "NK" = "#e72a8a", "Myeloid" = "#984ea3",
         "non-haematopoietic" = "#ff7f00", "T" = "#377eb8",
         "alpha-beta T cell" = "#377eb8", "B cell" = "#2f9e77",
         "mature B cell" = "#55c552", "not lymphocyte" = "#d95f02",
         "not mature B cell" = "#397037")

# dndscv palette
impact_colours <- c("Missense" = "#5f9fa0",
                    "Nonsense" = "#682c8b",
                    "Essential_Splice" = "#8e55a3",
                    "Indels" = "#d4773f",
                    "Synonymous" = "#b3b4b4",
                    "Non-coding" = "#606060")
```

# Samplesheet

We load the samplesheet.

```{r load_ss}
man_insp <-
  read_tsv("data/manual_inspection/2024-12-20_PD63118_PTA_BAF_LoH_CellType_Mut_Summary.tsv") %>%
  select(-c(id, run_id, seq_type)) %>%
  # fix loh_1p
  mutate(loh_1p = as.character(as.numeric(loh_1p))) %>%
  # synthesis celltype annotations
  mutate(
    celltype_VDJ_SHM = case_when(
      celltype_VDJ_recomb == "B cell" & celltype_SHM == "not mature B cell" ~ "non-mature B",
      celltype_VDJ_recomb == "B cell" & celltype_SHM == "mature B cell" ~ "mature B",
      celltype_VDJ_recomb == "alpha-beta T cell" ~ "T",
      celltype_VDJ_recomb == "not lymphocyte" ~ "non-lymphocyte",
      TRUE ~ celltype_VDJ_recomb) %>% factor(levels = c("mature B", "non-mature B", "T", "non-lymphocyte")))
ss <-
  read_csv("data/resolveome/samplesheet_local.csv") %>%
  # add manual inspection data
  left_join(man_insp) %>%
  # remove bad and uninspected cells
  filter(plate == 3, !chr_dropout, !suspected_doublet)
driver_genes <-
  readLines("../trencadis-seq/data/thyroid/driver_genes/driver_genes.txt")
```

# Somatic mutation genotyping

We load the genotyping and pile up reads across DNA and targeted DNA.

```{r load_vars}
genos <-
  ss %>%
  # get genotyping for dna and dnahyb
  filter(seq_type %in% c("dna", "dnahyb")) %>%
  distinct(seq_type, donor_id) %>%
  mutate(genos = paste0("out/nf-resolveome/", seq_type, "/", donor_id,
                        "/genotyping/mutations/", donor_id,
                        "_annotated_mutations.tsv")) %>%
  purrr::pmap(function(seq_type, donor_id, genos) {
    read_tsv(genos) %>%
      mutate(donor_id = donor_id) %>%
      # convert id -> cell_id
      inner_join(ss %>% select(id, cell_id)) %>%
      select(-id)
  }) %>%
  # pile up
  bind_rows() %>%
  select(-mut_vaf) %>%
  group_by(across(-matches("_depth$"))) %>%
  summarise(across(matches("_depth$"), \(x) sum(x, na.rm = TRUE))) %>%
  ungroup() %>%
  mutate(mut_vaf = mut_depth / total_depth) %>%
  # annotate mutations on 1p
  mutate(on_1p = ifelse(chr == "1" & pos < 123400000, TRUE, FALSE)) %>%
  # add manual inspection
  left_join(man_insp)

# return
genos
```

We plot the impact of the mutations genotyped.

```{r plot_impact}
genos %>%
  distinct(chr, pos, ref, alt, impact, gene) %>%
  add_count(gene) %>%
  ggplot(aes(x = reorder(gene, -n), fill = impact)) +
  geom_bar() +
  geom_text(aes(label = n, y = n), vjust = -0.5) +
  scale_fill_manual(values = impact_colours) +
  guides(x = guide_axis(angle = -90)) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 10)) +
  labs(x = "", y = "n mutations")
```

We generate a heatmap of all mutations.

```{r plot_heatmap, fig.width = 20, fig.height = 20}
plot_vaf_heatmap(genos, show_rownames = FALSE, p_title = "all mutations",
                 annotations = c("celltype_VDJ_SHM", "loh_1p"))
```

We generate a heatmap of the shared mutations.

```{r plot_heatmap_shared, fig.width = 20, fig.height = 20}
genos %>%
  filter(mut_depth > 0) %>%
  group_by(chr, pos, ref, alt) %>%
  filter(n() > 1) %>%
  plot_vaf_heatmap(show_rownames = FALSE, p_title = "shared mutations",
                   annotations = c("celltype_VDJ_SHM", "loh_1p"))
```

We filter to `mut_vaf` > 0.25 and `mut_depth` > 1.

```{r genos_filter}
genos_filter <-
  genos %>%
  filter(mut_vaf > 0.2, mut_depth > 1)

# return
genos_filter
```

```{r plot_heatmap_filtered, fig.width = 20, fig.height = 20}
plot_vaf_heatmap(genos_filter, p_title = "all mutations",
                 annotations = c("celltype_VDJ_SHM", "loh_1p"))
```

```{r plot_heatmap_filtered_shared, fig.width = 20, fig.height = 20}
genos_filter %>%
  group_by(chr, pos, ref, alt) %>%
  filter(n() > 1) %>%
  plot_vaf_heatmap(p_title = "shared mutations")
```

We generate a heatmap of the nonsynonymous mutations.

```{r plot_nonsynonymous}
genos_filter %>%
  filter(impact != "Synonymous") %>%
  plot_vaf_heatmap("non-synonymous mutations",
                   annotations = c("celltype_VDJ_SHM", "loh_1p"))
```

We generate a heatmap of the TNFRSF14 mutations.

```{r plot_heatmap_filtered_TNFRSF14, fig.width = 20, fig.height = 20}
genos_filter %>%
  filter(gene == "TNFRSF14", impact != "Synonymous") %>%
  plot_vaf_heatmap("non-synonymous TNFRSF14 mutations",
                   annotations = c("celltype_VDJ_SHM", "loh_1p"))
```

```{r plot_heatmap_filtered_TNFRSF14_nonsynonymous, fig.width = 20, fig.height = 20}
genos_filter %>%
  filter(gene == "TNFRSF14", impact != "Synonymous") %>%
  plot_vaf_heatmap("TNFRSF14 mutations")
```

We plot TNFRSF14 mutant calls in the 1p LOH and non-1p LOH cells.

```{r}
genos_filter %>%
  filter(gene == "TNFRSF14") %>%
  ggplot(aes(x = loh_1p, y = mut_vaf)) +
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(height = 0, width = 0.1) +
  theme_classic()
```

We plot the VAF distribution of the mutations.

```{r}
genos %>%
  left_join(ss) %>%
  filter(mut_vaf > 0) %>%
  ggplot(aes(x = mut_vaf, fill = on_1p & loh_1p)) +
  geom_histogram() +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set1")
genos_filter %>%
  left_join(ss) %>%
  ggplot(aes(x = mut_vaf, fill = on_1p & loh_1p)) +
  geom_histogram() +
  lims(x = c(0, NA)) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set1")
```

```{r plot_vaf_dist_hyb, fig.width = 20, fig.height = 20}
# % of B cells with a nonsynonymous mutation
n_mat_b_cells <-
  man_insp %>%
  filter(celltype_SHM == "mature B cell") %>%
  nrow()
p_dat <-
  genos_filter %>%
  filter(impact != "Synonymous") %>%
  group_by(gene) %>%
  summarise(n_cells = n_distinct(cell_id)) %>%
  mutate(prop_cells = n_cells / n_mat_b_cells)
p2 <-
  p_dat %>%
  ggplot(aes(x = reorder(gene, -prop_cells), y = prop_cells)) +
  geom_col(fill = "#d6630d") +
  theme_classic() +
  labs(x = "gene", y = "% of mature B cells with a driver mutation") +
  scale_y_continuous(expand = c(0, 0)) +
  guides(x = guide_axis(angle = -45))

# mutations per gene
p1 <-
  genos_filter %>%
  left_join(man_insp) %>%
  add_count(gene, name = "total") %>%
  count(gene, impact, total, celltype_SHM) %>%
  mutate(impact = ifelse(impact == "no-SNV", "Synonymous", impact) %>% factor(levels = rev(names(impact_colours)))) %>%
  left_join(p_dat %>% distinct(gene, prop_cells)) %>%
  ggplot(aes(x = reorder(gene, -prop_cells), y = n, fill = impact)) +
  geom_col() +
  scale_fill_manual(values = impact_colours) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = "n mutations")

# plot
p1 / p2
```

```{r plot_drivers_per_cell}
p_dat <-
  genos_filter %>%
  filter(impact != "Synonymous") %>%
  arrange(loh_1p) %>%
  mutate(cell_id = forcats::fct_inorder(cell_id)) %>%
  add_count(gene) %>%
  add_count(gene, cell_id, name = "n per cell") %>%
  mutate(gene = paste0(gene, " (", n, ")")) %>%
  arrange(n) %>%
  mutate(gene = forcats::fct_inorder(gene)) %>%
  distinct(cell_id, celltype_VDJ_SHM, gene, loh_1p, `n per cell`)

p1 <-
  p_dat %>%
  ggplot(aes(x = cell_id, y = 1, fill = loh_1p)) +
  geom_tile() +
  scale_fill_brewer(palette = "Set1") +
  theme_void()

p2 <-
  p_dat %>%
  ggplot(aes(x = cell_id, y = gene, fill = `n per cell`)) +
  geom_tile() +
  theme_classic() +
  guides(x = guide_axis(angle = -90)) +
  viridis::scale_fill_viridis()

p1 / p2 + plot_layout(heights = c(1, 10), guides = "collect")
```

# Somatic mutation calling

We load mutations generated by previous reports and we combine the calls.

```{r load_vafs}
# load calls
calls <-
  c("dna", "dnahyb") %>%
  purrr::set_names() %>%
  purrr::map(~
    system(paste0("ls reports/*_shared_muts_from_bj-somatic-variantcalling_",
           .x, "_filter_cells_cache/vafs_annotated_*.rds"), intern = TRUE) %>%
      readRDS()) %>%
  bind_rows() %>%
  select(-c(id, alt_vaf)) %>%
  # pile up
  group_by(across(-matches("_depth$"))) %>%
  summarise(across(matches("_depth$"), \(x) sum(x, na.rm = TRUE))) %>%
  ungroup() %>%
  mutate(alt_vaf = alt_depth / total_depth) %>%
  # remove common snp vars
  dplyr::filter(!common_snp) %>%
  # rename alt_* > mut_*
  rename_with(~ gsub("alt_", "mut_", .x), starts_with("alt_"))
```

Plot the distribution of mutation types per cell.

```{r plot_mut_types_all, fig.width = 12}
# all muts
calls %>%
  filter(!is.na(impact)) %>%
  mutate(group = "all") %>%
  bind_rows(calls %>% filter(impact != "Non-coding") %>% mutate(group = "coding")) %>%
  add_count(group, cell_id) %>%
  ggplot(aes(x = reorder(cell_id, -n), fill = impact)) +
  geom_bar() +
  scale_fill_manual(values = impact_colours) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  guides(x = guide_axis(angle = -90)) +
  labs(x = "") +
  facet_grid(group ~ ., scales = "free_y")
```

# VAF distribution

Plot the VAF distribution.

```{r plot_vaf_dist, fig.height = 4, fig.width = 5}
c(1, 5, 10, 20, 50) %>%
  purrr::walk(function(min_total_depth) {
    p_dat <- calls %>% filter(total_depth >= min_total_depth)
    p <-
      p_dat %>%
      ggplot(aes(x = mut_vaf)) +
      geom_histogram() +
      theme_classic() +
      labs(title = paste("VAF distribution, min depth =", min_total_depth),
           subtitle = paste(nrow(p_dat), "mutations"), x = "alt VAF")
    print(p)
  })
```