---
title: "Shared mutations from the bj-somatic-variantcalling WGS output"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  rerun: false
---

```{r setup, include = F}
# bsub -q basement -M50000 -R 'span[hosts=1] select[mem>50000] rusage[mem=50000]' -J plot_shared_muts_from_bj-somatic-variantcalling_wgs -o log/%J_plot_shared_muts_from_bj-somatic-variantcalling_wgs.out -e log/%J_plot_shared_muts_from_bj-somatic-variantcalling_wgs.err Rscript -e "rmarkdown::render('reports/20250304_plot_shared_muts_from_bj-somatic-variantcalling_wgs.Rmd', output_file = '20250304_plot_shared_muts_from_bj-somatic-variantcalling_wgs.html', output_dir = 'reports/')"
# rmarkdown::render('reports/20250304_plot_shared_muts_from_bj-somatic-variantcalling_wgs.Rmd', output_file = '20250304_plot_shared_muts_from_bj-somatic-variantcalling_wgs.html', output_dir = 'reports/')

# chunk options
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE,
                      dpi = 600,
                      message = FALSE,
                      cache.path = "reports/20250304_plot_shared_muts_from_bj-somatic-variantcalling_wgs_cache/")

# set buffer for reading long lines in vcfs
Sys.setenv("VROOM_CONNECTION_SIZE" = 2^20) # 1MB buffer

# libraries
library(magrittr)
library(ggplot2)
library(dndscv)
library(GenomicRanges)
library(ggtree)

# palettes
impact_colours <- c("Essential_Splice" = "#8e55a3",
                    "Nonsense" = "#682c8b",
                    "Missense" = "#5f9fa0",
                    "Indels" = "#d4773f",
                    "Synonymous" = "#b3b4b4",
                    "Non-coding" = "#606060")

# function: get the trinucleotide context of the muts
get_mut_trinucs <- function(dat) {
  dat %>%
    # get snps only
    dplyr::filter(nchar(ref) == 1, nchar(alt) == 1) %>%
    # get the trinucleotide context
    dplyr::rowwise() %>%
    dplyr::mutate(
      trinuc_ref = Rsamtools::scanFa("data/gatk/grch38/genome.fa",
                                     GRanges(chr, IRanges(pos - 1,
                                                          pos + 1))) %>%
                    as.vector()) %>%
    dplyr::ungroup() %>%
    # annotate the mutation from the pyrimidine base
    dplyr::mutate(
      sub = paste(ref, alt, sep = ">"),
      sub_py = dplyr::case_when(
        ref %in% c("A", "G") ~ chartr("TCGA", "AGCT", sub),
        TRUE ~ sub),
      trinuc_ref_py = dplyr::case_when(
        ref %in% c("A", "G") ~ chartr("TCGA", "AGCT", trinuc_ref),
        TRUE ~ trinuc_ref))
}

# function: plot the trinucleotide context of the muts
plot_mut_trinucs <- function(p_dat, p_title = "") {
  sub_colours <-
    c("C>A" = "dodgerblue", "C>G" = "black", "C>T" = "red",
      "T>A" = "grey70", "T>C" = "olivedrab3", "T>G" = "plum2")
  nucs <- c("T", "C", "G", "A")
  py_nucs <- c("C", "T")
  all_subs <-
    expand.grid(do.call(paste0, expand.grid(nucs, py_nucs, nucs)), nucs) %>%
    dplyr::transmute(trinuc_ref_py = Var1, ref = substr(Var1, 2, 2), alt = Var2,
                     sub_py = paste0(ref, ">", alt)) %>%
    dplyr::filter(alt != ref)

  p_dat %>%
    dplyr::mutate(n_cells = dplyr::n_distinct(id)) %>%
    dplyr::add_count(mut_id, name = "n_cells_w_mut") %>%
    dplyr::count(sub_py, trinuc_ref_py) %>%
    dplyr::right_join(all_subs) %>%
    dplyr::mutate(n = ifelse(is.na(n), 0, n)) %>%
    ggplot(aes(x = trinuc_ref_py, y = n, fill = sub_py)) +
    geom_col() +
    facet_grid(~ sub_py, scales = "free_x") +
    guides(x = guide_axis(angle = 90)) +
    theme_minimal() +
    theme(axis.text.x = element_text(family = "mono"),
          legend.position = "none") +
    scale_fill_manual(values = sub_colours) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = p_title, subtitle = paste(nrow(p_dat), "mutations"),
         x = "", y = "number of mutations")
}

# function: plot VAF heatmap
plot_vaf_heatmap <- function(p_dat, p_title, show_rownames = FALSE) {
  # reshape data for heatmap
  heatmap_data <-
    p_dat %>%
    dplyr::add_count(mut_id, name = "n_cells_w_mut") %>%
    reshape2::dcast(mut_id + n_cells_w_mut ~ id, value.var = "alt_vaf")
  rownames(heatmap_data) <- heatmap_data$mut_id
  heatmap_matrix <- as.matrix(heatmap_data[, -c(1:3)]) # drop annotations
  n_total_muts <- dplyr::n_distinct(p_dat$mut_id)
  n_total_cells <- dplyr::n_distinct(p_dat$id)

  # calculate the number of mutations per column (id)
  annotation_col <-
    p_dat %>%
    dplyr::count(id, run_id, celltype, qual, loh_1p, name = "n_muts") %>%
    dplyr::mutate(run_id = as.character(run_id)) %>%
    tibble::column_to_rownames("id")

  # replace NAs with 0
  heatmap_matrix[is.na(heatmap_matrix)] <- 0

  # prepare annotations for columns
  annotation_row <- data.frame(n = heatmap_data$n)
  rownames(annotation_row) <- heatmap_data$mut_id

  # plot
  pheatmap::pheatmap(
    mat = heatmap_matrix,
    cluster_rows = ifelse(nrow(heatmap_matrix) < 65536, TRUE, FALSE),
    cluster_cols = TRUE,
    show_rownames = show_rownames,
    annotation_row = annotation_row,
    annotation_col = annotation_col,
    color = colorRampPalette(c("white", "blue"))(50),
    main = paste("VAF heatmap:", p_title, "\n",
                 n_total_cells, "cells,", n_total_muts, "mutations"))
}
```

We pick out a BaseJumper run.

```{r bj_dir}
# get the latest run
bj_dir <- "out/BaseJumper/bj-somatic-variantcalling/wgs/PD63118_250306_232001/"
vcfs_subdir <- "/SOMATIC_VARIANT_WORKFLOW_Heuristic_Filter_SUBSET_VCF_VARIANTS/"
```

We load the samplesheet.

```{r load_data}
man_insp <-
  readr::read_tsv("data/manual_inspection/2024-12-20_PD63118_PTA_BAF_LoH_CellType_Mut_Summary.tsv") %>%
  dplyr::mutate(run_id = as.character(run_id))
metadata <-
  readr::read_csv("data/resolveome/samplesheet_local.csv") %>%
  dplyr::left_join(man_insp) %>%
  dplyr::mutate(
    # define celltypes
    celltype = dplyr::case_when(
      celltype_VDJ_recomb == "alpha-beta T cell" ~ "T cell",
      celltype_VDJ_recomb == "B cell" & celltype_SHM == "mature B cell" ~
        "mature B cell",
      celltype_VDJ_recomb == "B cell" & celltype_SHM == "not mature B cell" ~
        "non-mature B cell",
      TRUE ~ "uncertain"),
    # define quality (good vs bad cells)
    qual = ifelse(!chr_dropout & !suspected_doublet, "good", "bad"),
    # define loh_1p
    loh_1p = dplyr::case_when(loh_1p ~ "1p LOH", !loh_1p ~ "no 1p LOH",
                              TRUE ~ "unknown"))
ss <-
  readr::read_csv("out/BaseJumper/bj-somatic-variantcalling/wgs/samplesheet.csv")
```

We load the driver genes.

```{r driver_genes}
driver_genes <-
  readLines("../trencadis-seq/data/thyroid/driver_genes/driver_genes.txt")
```

# `MultiQC`

We load the `multiqc` output.

```{r multiqc, fig.width = 8}
multiqc <-
  paste0(bj_dir, "/multiqc/multiqc_data/multiqc_PROJECT ID PLOT_TITLE_1.txt") %>%
  readr::read_tsv() %>%
  dplyr::rename(id = Sample) %>%
  dplyr::left_join(metadata %>% dplyr::select(-bam)) 

multiqc %>%
  # clean
  dplyr::filter(!chr_dropout, !suspected_doublet) %>%
  ggplot(aes(x = reorder(id, -MEAN_COVERAGE), y = MEAN_COVERAGE)) +
  geom_col() +
  geom_errorbar(aes(ymin = MEAN_COVERAGE - SD_COVERAGE,
                    ymax = MEAN_COVERAGE + SD_COVERAGE),
                width = 0.2) +
  guides(x = guide_axis(angle = -90)) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "", y = "mean coverage")

multiqc %>%
  dplyr::left_join(metadata) %>%
  dplyr::mutate(chr_dropout = ifelse(chr_dropout, "chr dropout",
                                     "no chr dropout")) %>%
  ggplot(aes(x = reorder(id, -HET_SNP_SENSITIVITY), y = HET_SNP_SENSITIVITY)) +
  geom_col() +
  facet_grid(~ chr_dropout, scales = "free_x", space = "free_x") +
  guides(x = guide_axis(angle = -90)) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "", y = "heterozygous SNP sensitivity")
```

# `BaseJumper` somatic variant calls

We load the variants and add cell-level metadata.

```{r load_vars}
# get vafs, add cell-level metadata
vafs <- xfun::cache_rds({
  ss$biosampleName %>%
    purrr::set_names() %>%
    purrr::map(function(id) {
      vcf_file <-
        paste0(bj_dir, vcfs_subdir, id, "_somatic_filtered_variants.vcf.gz")
      if (file.exists(vcf_file)) {
        vcf <-
          readr::read_tsv(vcf_file, comment = "##") %>%
          {split(., .$FORMAT)} %>%
          purrr::map(function(vcf_i) {
            format_i <- stringr::str_split_1(unique(vcf_i$FORMAT), ":")
            vcf_i %>%
              dplyr::mutate(gt = get(id)) %>%
              tidyr::separate_wider_delim("gt", delim = ":", names = format_i) %>%
              janitor::clean_names() %>%
              dplyr::mutate(chr = number_chrom, total_depth = dp,
                            allele = paste(ref, alt, sep = ","),
                            alt = gsub(",.*", "", alt)) %>%
              tidyr::separate_longer_delim(cols = c("allele", "ad"),
                                           delim = ",") %>%
              dplyr::select(chr, pos, ref, alt, allele, total_depth, ad) %>%
              dplyr::mutate(name = dplyr::case_when(allele == ref ~ "ref_depth",
                                                    allele == alt ~ "alt_depth",
                                                    TRUE ~ "other_depth")) %>%
              dplyr::filter(name != "other_depth") %>%
              tidyr::pivot_wider(
                names_from = "name", values_from = "ad",
                id_cols = c("chr", "pos", "ref", "alt", "total_depth")) %>%
              readr::type_convert() %>%
              dplyr::mutate(alt_vaf = alt_depth / total_depth)
          }) %>%
          dplyr::bind_rows()
      }
    }) %>%
    purrr::compact() %>%
    dplyr::bind_rows(.id = "id") %>%
    # annotate celltypes
    dplyr::left_join(metadata)
  }, file = "vafs.rds", rerun = params$rerun)
```

# Variant annotation

We annotate the variants using `dndscv`.

```{r annot_muts}
refdb_file <-
  "../reference/dndscv/RefCDS_human_GRCh38_GencodeV18_recommended.rda"
vafs <-
  xfun::cache_rds({
    vafs %>%
      dplyr::transmute(sampleID = id, chr = gsub("chr", "", chr), pos, ref,
                       mut = alt) %>%
      dplyr::distinct() %>%
      dndscv(refdb = refdb_file, outp = 1,
             max_coding_muts_per_sample = Inf,
             max_muts_per_gene_per_sample = Inf) %>%
      {.$annotmuts} %>%
      dplyr::transmute(chr = paste0("chr", chr), pos, ref, alt = mut, gene, pid,
                       id = sampleID, aachange, ntchange, codonsub, impact) %>%
      dplyr::full_join(vafs) %>%
      dplyr::mutate(impact = ifelse(is.na(impact), "Non-coding", impact) %>%
                      factor(levels = rev(names(impact_colours))),
                    mut_id = paste(chr, pos, ref, alt, sep = "-"))
  }, file = "vafs_annotated.rds", rerun = params$rerun)
```

# Cell filtering

We mark cells with chromosomal dropout and suspected doublets.

```{r filter_cells}
vafs <-
  xfun::cache_rds({
    vafs %>%
      # mark cells with chr dropout and suspected doublets
      dplyr::mutate(qual = ifelse(!chr_dropout & !suspected_doublet,
                                  "good", "bad"))
  }, file = "vafs_w_filter.rds", rerun = params$rerun)
```

# Mutation types per cell

Plot the distribution of mutation types per cell.

```{r plot_mut_types_all, fig.width = 12}
# all muts
vafs %>%
  dplyr::add_count(id) %>%
  ggplot(aes(x = reorder(id, -n), fill = impact)) +
  geom_bar() +
  scale_fill_manual(values = impact_colours) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  guides(x = guide_axis(angle = -90)) +
  labs(x = "") +
  facet_grid(~ qual, scales = "free_x", space = "free_x")
```

```{r plot_mut_types_coding, fig.width = 12}
# coding muts
vafs %>%
  dplyr::filter(impact != "Non-coding") %>%
  dplyr::add_count(id) %>%
  ggplot(aes(x = reorder(id, -n), fill = impact)) +
  geom_bar() +
  scale_fill_manual(values = impact_colours) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  guides(x = guide_axis(angle = -90)) +
  labs(x = "") +
  facet_grid(~ qual, scales = "free_x", space = "free_x")
```

# Mutation spectra

Get the mutational spectra in the good cells.

```{r get_spectra}
# get trinucleotide context
vafs_trinuc <-
  xfun::cache_rds({
    vafs %>%
      dplyr::filter(qual == "good") %>%
      get_mut_trinucs()
  }, file = "vafs_trinuc.rds", rerun = params$rerun)
```

## Plot trinucleotide spectrum (good cells)

```{r plot_spectra_good, fig.height = 5, fig.width = 11}
plot_mut_trinucs(vafs_trinuc, "good cells")
```

## Plot trinucleotide spectrum per celltype (good cells)

```{r plot_spectra_good_per_ct, fig.height = 5, fig.width = 11}
split(vafs_trinuc, vafs_trinuc$celltype) %>%
  purrr::walk(function(vafs_trinuc_i) {
    print(plot_mut_trinucs(vafs_trinuc_i, unique(vafs_trinuc_i$celltype)))
  })
```

## Plot trinucleotide spectrum per cell (good cells)

```{r plot_spectra_per_cell, fig.height = 5, fig.width = 11}
split(vafs_trinuc, vafs_trinuc$id) %>%
  purrr::walk(function(vafs_trinuc_i) {
    print(plot_mut_trinucs(vafs_trinuc_i, unique(vafs_trinuc_i$id)))
  })
```

# VAF distribution

Plot the VAF distribution.

```{r plot_vaf_dist, fig.height = 8, fig.width = 5}
c(1, 5, 10, 20, 50) %>%
  purrr::walk(function(min_total_depth) {
    p_dat <- vafs %>% dplyr::filter(total_depth >= min_total_depth)
    p <-
      p_dat %>%
      ggplot(aes(x = alt_vaf)) +
      geom_histogram() +
      theme_classic() +
      labs(title = paste("VAF distribution, min depth =", min_total_depth),
           subtitle = paste(nrow(p_dat), "mutations"), x = "alt VAF") +
      facet_grid(qual ~ .)
    print(p)
  })
```

# VAF heatmap

## Plot all mutations (all cells)

```{r plot_vaf_heatmap, fig.height = 8, fig.width = 15}
p <- plot_vaf_heatmap(vafs, "all muts")
print(p)
```

## Plot all shared mutations (all cells)

```{r plot_vaf_heatmap_shared, fig.height = 8, fig.width = 15}
p_dat_shared <-
  vafs %>%
  dplyr::mutate(n_cells = dplyr::n_distinct(id)) %>%
  dplyr::add_count(mut_id, name = "n_cells_w_mut") %>%
  dplyr::mutate(prop_cells_w_mut = n_cells_w_mut / n_cells) %>%
  # shared mutations
  dplyr::filter(n_cells_w_mut > 1)
p <- plot_vaf_heatmap(p_dat_shared, "all shared muts")
print(p)
```

## Plot all shared mutations (good cells)

```{r plot_vaf_heatmap_shared_good, fig.height = 8, fig.width = 8}
p_dat_shared_good <- p_dat_shared %>% dplyr::filter(qual == "good")
p <- plot_vaf_heatmap(p_dat_shared_good, "all shared muts (good cells)")
print(p)
```

## Plot all shared mutations (good cells, % cells w mut < 0.3)

In order to exclude some of the germline noise, we look just at mutations 
present in <30% of cells.

```{r plot_vaf_heatmap_shared_good_filtered, fig.height = 8, fig.width = 8}
p_dat_shared_filtered <-
  p_dat_shared_good %>%
  dplyr::filter(prop_cells_w_mut < 0.3)
p <- plot_vaf_heatmap(p_dat_shared_filtered,
                      "shared muts (good cells, % cells w mut < 0.3)")
print(p)
```

## Plot coding shared mutations (all cells)

```{r plot_vaf_heatmap_shared_coding, fig.height = 8, fig.width = 15}
# look at coding muts
p_dat_coding <-
  p_dat_shared %>%
  dplyr::filter(impact != "Non-coding")
p <- plot_vaf_heatmap(p_dat_coding, "shared coding muts (all cells)", TRUE)
print(p)
```

# Case study: `plate3_wellD4` & `plate3_wellE11`

We see that cells `plate3_wellD4` and `plate3_wellE11` share a block of 
mutations. We will look at the spectra of these mutations that are exclusively 
shared between these two cells.

```{r plot_spectra_shared_exclusive, fig.height = 5, fig.width = 11}
excl_vafs_trinuc <-
  vafs %>%
  dplyr::group_by(chr, pos, ref, alt) %>%
  dplyr::filter(paste(sort(unique(id)), collapse = ",") == "plate3_wellD4_dna_run49882,plate3_wellE11_dna_run49882") %>%
  get_mut_trinucs()
plot_mut_trinucs(excl_vafs_trinuc,
                 "plate3_wellD4, plate3_wellE11 exclusive shared muts")
```

# Sequoia

We load the tree.

```{r sequoia}
tree <-
  file.path(bj_dir, "/SOMATIC_VARIANT_WORKFLOW_Heuristic_Filter_SEQUOIA/") %>%
  list.files(pattern = "tree_with_branch_length.tree", full.names = TRUE) %>%
  ape::read.tree()
```

## Phylogeny (all cells)

```{r plot_tree, fig.height = 10}
# colour clonal tipes
clones <-
  list(
    clone1 = c("plate3_wellD4_dna_run49882", "plate3_wellE11_dna_run49882"),
    clone2 = c("plate3_wellD11_dna_run49882", "plate3_wellA10_dna_run49882"),
    clone3 = c("plate3_wellD3_dna_run49882", "plate3_wellB7_dna_run49882"),
    clone4 = c("plate3_wellE8_dna_run49882", "plate3_wellF10_dna_run49882"))
tip_col <- rep("black", length(tree$tip.label)) %>% setNames(tree$tip.label)
tip_pal <- RColorBrewer::brewer.pal(length(clones), "Dark2")
purrr::walk2(clones, tip_pal, function(clone, col) {
  tip_col[clone] <<- col
})

# colour celltypes
tip_annotations <-
  metadata %>%
  dplyr::filter(id %in% tree$tip.label) %>%
  dplyr::mutate(celltype = ifelse(is.na(celltype), "other", celltype)) %>% 
  {split(.$id, .$celltype)}
tip_annotations <- tip_annotations[tree$tip.label]

groupOTU(ggtree(tree), tip_annotations, "celltype") +
  geom_tiplab(show.legend = FALSE) +
  aes(colour = celltype) +
  scale_colour_brewer(palette = "Set2")

# plot tree with colored tip labels
plot(tree, tip.color = tip_col, cex = 0.6)
ape::axisPhylo(side = 1, backward = FALSE)
```

## Phylogeny (good cells)

```{r plot_tree_good, fig.height = 8}
good_ids <- dplyr::filter(vafs, qual == "good")$id %>% unique()
good_tree <- ape::drop.tip(tree, setdiff(tree$tip.label, good_ids))
tip_col <-
  rep("black", length(good_tree$tip.label)) %>%
  setNames(good_tree$tip.label)
purrr::walk2(clones, tip_pal, function(clone, col) {
  tip_col[clone] <<- col
})

plot(good_tree, tip.color = tip_col, cex = 0.6)
ape::axisPhylo(side = 1, backward = FALSE)
```

# Driver mutations

```{r drivers}
vafs %>%
  dplyr::filter(gene %in% driver_genes) %>%
  dplyr::select(mut_id, gene, id, aachange, ntchange, codonsub, impact,
                ref_depth, alt_depth, alt_vaf) %>%
  knitr::kable()
```

# Beta binomial

```{r beta_binom, eval = FALSE}
# get nv and nr
mats <-
  list("NR", "NV") %>%
  purrr::set_names() %>%
  purrr::map(function(mat_type) {
    file.path(bj_dir, "SOMATIC_VARIANT_WORKFLOW_Heuristic_Filter_SEQUOIA") %>%
      list.files(pattern = paste0("Mat_", mat_type), full.names = TRUE) %>%
      readr::read_tsv(col_names = FALSE) %>%
      tibble::column_to_rownames("X1") %>%
      as.matrix()
  })

# estimate overdispersion
estimateRho_gridml <- function(NV_vec, NR_vec) {
  # rho will be bounded within 1e-6 and 0.89
 rhovec <- 10^seq(-6, -0.05, by = 0.05)
 mu <- sum(NV_vec) / sum(NR_vec)
 ll <- sapply(rhovec, function(rhoj) sum(dbetabinom(x = NV_vec, size = NR_vec, rho = rhoj, prob = mu, log = TRUE)))
 return(rhovec[ll == max(ll)][1])
}

rho_est <- pvalue <- rep(NA, nrow(mats$NR))
require(VGAM)
for (k in 1:nrow(mats$NR)) {
	rho_est[k] <- estimateRho_gridml(NV_vec = as.numeric(mats$NV[k,]), NR_vec = as.numeric(mats$NR[k,]))
	if (k%%1000 == 0) {
		print(k)
	}
}

# num_samples: those with at least 2 supporting reads
df <-
  data.frame(pos = 1:length(rho_est),
             rho_est = log(rho_est),
             num_samples = rowSums(NV[rownames(NV), ] > 1))

ggplot(df, aes(x = pos, y = rho_est)) +
  geom_point(aes(colour = factor(num_samples), alpha = 0.5)) +
  theme(legend.position = "none")

hist(df$num_samples, breaks = 20, xlab = "Number of samples with mutation",
     ylab = "Number of mutations", main = "pmin(20)")

df$global_vaf <-
  apply(alts[rownames(vafs),], 1, sum) / apply(covs[rownames(vafs), ], 1, sum)
ggplot(df, aes(x = pos, y = rho_est)) +
  geom_point(aes(colour = global_vaf))
```