---
title: "Coverage and genotyping in PTA"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include = F, message = F, warning = F, echo = F}
# rmarkdown::render('src/05c_plot_nf-resolveome.Rmd', output_file = 'report_run_49686_vs_49882.html', output_dir = 'reports/')

# chunk options
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE,
                      dpi = 300,
                      message = FALSE)

# libraries
library(magrittr)
library(ggplot2)

# pattient
curr_pdid <- "PD63118"

# samplesheet
ss <-
  system("ls out/nf-resolveome/*/samplesheet.tsv", intern = TRUE) %>%
  purrr::map(readr::read_tsv) %>%
  dplyr::bind_rows()
```

Here, we are analysing the mutations genotyped in PTA data. We want to 
investigate genotyping efficiency, VAF distribution, and allelic imbalance.

### Coverage

Coverage was calculated using `MOSDEPTH` across the genome and across the immune 
panel for each sample.

```{r load_mosdepth}
# get summary
summ <-
  ss %>%
  purrr::pmap(function(donor_id, id, run, plex_n, ...) {
    paste0("out/nf-resolveome/", run, "/", donor_id, "/", id,
            "/mosdepth/", id, ".mosdepth.summary.txt") %>%
      readr::read_tsv() %>%
      dplyr::mutate(id = id, run = run, donor_id = donor_id, plex_n = plex_n) %>%
      # remove extra chromosomes
      dplyr::filter(!grepl("^GL", chrom), !grepl("Un", chrom),
                    !grepl("random|HLA", chrom), !grepl("alt$", chrom))
  }) %>%
  dplyr::bind_rows() 

# get cov
cov <-
  ss %>%
  purrr::pmap(function(donor_id, id, run, plex_n, ...) {
    c("region", "global") %>%
      purrr::set_names() %>%
      purrr::map(function(lvl) {
        readr::read_tsv(paste0("out/nf-resolveome/", run, "/", donor_id, "/", id,
                               "/mosdepth/", id, ".mosdepth.", lvl, ".dist.txt"),
                        col_names = c("chr", "cov", "prop")) %>%
          # remove extra chromosomes
          dplyr::filter(!grepl("^GL", chr), !grepl("Un", chr),
                        !grepl("random|HLA", chr), !grepl("alt$", chr),
                        prop >= 0.01)
      }) %>%
      dplyr::bind_rows(.id = "lvl") %>%
      dplyr::mutate(id = id, run = run, donor_id = donor_id, plex_n = plex_n)
  }) %>%
  dplyr::bind_rows()
```

We can plot the mean coverage...

```{r mean_cov, fig.width = 10}
summ %>%
  dplyr::filter(chrom %in% c("total", "total_region")) %>%
  dplyr::mutate(lvl = ifelse(grepl("region$", chrom), "region", "global"),
                chrom = gsub("_.*", "", chrom)) %>%
  ggplot(aes(x = tidytext::reorder_within(id, -mean, run), y = mean, fill = lvl)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Mean coverage",
       x = "Plex",
       y = "Mean coverage") +
  ggh4x::facet_grid2(~ run, scales ="free_x", space="free_x") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0)) +
  tidytext::scale_x_reordered()
```

We can plot the coverage distribution...

```{r cov_dist}
cov %>%
  dplyr::filter(chr == "total") %>%
  ggplot(aes(x = cov, y = prop, color = id, group = id)) +
  geom_line() +
  labs(title = paste("Coverage distribution"),
       x = "Coverage",
       y = "Proportion of bases at coverage") +
  theme_minimal() +
  facet_grid(run ~ lvl) +
  theme(legend.position = "none")
```

### Genotyping

For patient `r curr_pdid`, we have genotyped mutations from two sources:

- Mutations from targeted immune and whole exome Nanoseq

- Heterozygous, common SNPs from `CaVEMan` output

We load the mutations from Nanoseq and the SNPs from CaVEMan...

```{r load_mutations}
muts_and_snps <-
  readr::read_tsv("out/nf-resolveome/mutations.tsv") %>%
  dplyr::distinct(source, pdid, chr, pos, ref, mut)
muts_and_snps %>% dplyr::count(source)
```

Next, we load genotyping outputs and calculate VAFs...

```{r load_geno}
mut_depth_bins <- c("0", "1", ">1", ">5", ">10", ">50")
geno <-
  ss %>%
  purrr::pmap(function(donor_id, id, run, plex_n, ...) {
    paste0("out/nf-resolveome/", run, "/", donor_id, "/", id,
            "/genotyping/", id, "_genotyped_mutations.tsv") %>%
      readr::read_tsv(col_types = list(chr = readr::col_character())) %>%
      dplyr::mutate(id = id, run = run, donor_id = donor_id, plex_n = plex_n)
  }) %>%
  dplyr::bind_rows() %>%
  dplyr::mutate(mut_vaf = mut_depth / total_depth,
                mut_id = paste(chr, pos, ref, mut)) %>%
  dplyr::left_join(muts_and_snps) %>%
  # bin mutant depths
  dplyr::mutate(mut_depth_bin = dplyr::case_when(mut_depth > 50 ~ ">50",
                                                 mut_depth > 10 ~ ">10",
                                                 mut_depth > 5 ~ ">5",
                                                 mut_depth > 1 ~ ">1",
                                                 mut_depth == 1 ~ "1",
                                                 mut_depth == 0 ~ "0") %>%
                    factor(levels = mut_depth_bins))
```

We now plot the VAF distributions...

```{r plot_vaf, fig.width = 10, fig.height = 8}
# function: plot VAF distribution
plot_vaf_dist <- function(p_dat) {
  p_dat %>%
    ggplot(aes(x = mut_vaf, fill = mut_depth_bin, colour = mut_depth_bin)) +
    geom_histogram(bins = 100) +
    facet_grid(source ~ run, scales = "free_y") +
    theme_minimal() +
    viridis::scale_fill_viridis(discrete = TRUE) +
    viridis::scale_color_viridis(discrete = TRUE)
}

# plot vaf dist
plot_vaf_dist(geno) +
  ggtitle("Distribution of VAFs")

# plot non-zero vaf dist
plot_vaf_dist(dplyr::filter(geno, mut_depth > 0)) +
  ggtitle("Distribution of non-zero VAFs")

# plot vaf dist with increasing mut depth
seq(1,10) %>%
  purrr::map(function(min_mut_depth) {
    geno %>%
      dplyr::filter(mut_depth >= min_mut_depth) %>%
      plot_vaf_dist() +
      ggtitle(paste("Distribution of VAFs, min", min_mut_depth, "mut depth"))
  })
```

How many mutations and SNPs were genotyped per cell?

```{r n_genotyped, fig.width = 10, fig.height = 8}
geno %>%
  dplyr::mutate(genotyped = ifelse(mut_depth > 0, mut_id, NA)) %>%
  dplyr::group_by(source, id, run) %>%
  dplyr::mutate(n = dplyr::n_distinct(mut_id, na.rm = TRUE)) %>%
  dplyr::group_by(source, id, run, n, mut_depth_bin) %>%
  dplyr::summarise(n_genotyped = dplyr::n_distinct(genotyped, na.rm = TRUE)) %>%
  dplyr::mutate(prop_genotyped = n_genotyped / n) %>%
  dplyr::group_by(run, id) %>%
  dplyr::mutate(total_prop = sum(prop_genotyped)) %>%
  ggplot(aes(x = tidytext::reorder_within(id, -total_prop, run),
             y = prop_genotyped, fill = mut_depth_bin)) +
  geom_col(position = "stack") +
  ggh4x::facet_grid2(source ~ run, scales = "free", space = "free_x") +
  tidytext::scale_x_reordered(guide = guide_axis(angle = -90)) +
  labs(title = "Proportion of mutations genotyped per source per mut depth",
       x = "Plex",
       y = "Proportion of mutations genotyped") +
  theme_minimal() +
  viridis::scale_fill_viridis(discrete = TRUE)
```

Which mutations had the top VAFs and what were their depths?

```{r top_vafs}
geno %>%
  dplyr::filter(source == "nanoseq_mutations") %>%
  dplyr::arrange(-mut_vaf, -mut_depth) %>%
  dplyr::filter(mut_vaf > 0) %>%
  dplyr::select(source, id, mut_id, mut_vaf, mut_depth, total_depth) %>%
  head(15)
```

How many mutations were shared between multiple cells?

```{r shared_muts}
p_dat <-
  geno %>%
  dplyr::filter(mut_depth > 0) %>%
  dplyr::group_by(mut_id, source) %>%
  dplyr::summarise(n_samples_w_mut = dplyr::n_distinct(id),
                   mean_mut_vaf = mean(mut_vaf),
                   ids = id %>% unique() %>% sort() %>% paste(collapse = ",")) %>%
  dplyr::ungroup() %>%
  dplyr::count(source, n_samples_w_mut) %>%
  dplyr::arrange(-n_samples_w_mut) %>%
  {split(., .$source)}

p_dat %>%
  dplyr::bind_rows() %>%
  ggplot(aes(x = n_samples_w_mut, y = n)) +
  geom_col() +
  facet_wrap(~ source)
```

We can compare the unbiased NanoSeq VAF to the PTA VAF. The axes in this plot 
are log-scaled. The dashed line represents a VAF of 0.5.

```{r nanoseq_vaf_vs_pta_vaf}
nanoseq_muts <-
  readr::read_tsv("data/nanoseq/samplesheet.tsv") %>%
  purrr::pmap(function(muts_file, seq_type) {
    readr::read_tsv(muts_file) %>%
      dplyr::mutate(seq_type = seq_type)
  }) %>%
  dplyr::bind_rows() %>%
  dplyr::mutate(pdid = stringr::str_sub(sampleID, 1, 7)) %>%
  dplyr::filter(pdid == curr_pdid) %>%
  dplyr::mutate(mut_id = paste(chr, pos, ref, mut))
geno %>%
  dplyr::inner_join(nanoseq_muts) %>%
  dplyr::filter(total_depth > 0) %>%
  dplyr::mutate(unbiased_vaf = (bam_mut - times_called) / (bam_cov - duplex_cov)) %>%
  ggplot(aes(x = unbiased_vaf, y = mut_vaf, size = total_depth, alpha = total_depth)) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_point() +
  labs(main = "PTA VAF versus NanoSeq unbiased VAF",
       x = "NanoSeq unbiased VAF",
       y = "PTA VAF") +
  theme_classic()
```

