---
title: "VDJ coverage"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
    css: "styles.css" # Link to your custom CSS file
---


```{r setup, include = F, message = F, warning = F, echo = F}
# rmarkdown::render('src/08d_knit_VDJ_coverage.Rmd', output_file = 'VDJ_coverage.html', output_dir = 'out/vdj_reconstruction/')

# chunk options
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE,
                      dpi = 300,
                      message = FALSE)

# libraries
library(magrittr)
```

```{r plots, results = 'asis', echo = F}
# dirs
wd <- getwd()
plots <-
  system(paste0("ls ", wd,
                "/out/vdj_reconstruction/PD63118_49882/plex*/plots/plex*.png"),
         intern = TRUE) %>%
  tibble::tibble(plot = .) %>%
  dplyr::mutate(name = basename(plot) %>% tools::file_path_sans_ext()) %>%
  tidyr::separate_wider_delim(name, delim = "_", cols_remove = FALSE,
                              names = c("id", "chr", "class", "type"),
                              too_many = "merge") %>%
  dplyr::mutate(plex_n = as.numeric(gsub("plex", "", id)),
                id = id %>% forcats::fct_reorder(plex_n)) %>%
  dplyr::arrange(plex_n) %>%
  {split(., .$id)} %>%
  purrr::map(~ split(.x, .x$class))

# header params
knit_header <-
  '```{r, echo = F, include = T, message = T, fig.align = "center", fig.pos = "H", out.width = "1200px"}'

# initialise chunks
chunks <- c()

# loop through plots
purrr::walk2(names(plots), plots, function(id_i, id_dat) {
  chunks <<- c(chunks, paste0("## ", id_i))
  purrr::walk2(names(id_dat), id_dat, function(class_i, class_dat) {
    chunks <<- c(chunks, paste0("### ", class_i))
    purrr::walk(class_dat$plot, function(dat) {
      knit_plot <- c(
        '', knit_header,
        paste0('knitr::include_graphics("', dat, '")'),
        '```', '')
      knitted_plot <-
        knitr::knit_child(text = knit_plot, envir = environment(),
                          quiet = TRUE)
      chunks <<- c(chunks, knitted_plot)
    })
  })
})

# return
cat(unlist(chunks), sep = "\n")
```